### 2013-11-09 src2/thesis.tex
update

\section*{Abstract}
Abstract abstract

@@

@@

Computers today are increasingly parallel. Each new generation of general purpose processors
(CPUs) and of graphics processors (GPUs), brings more parallelism. The question 
of how to program these computers most efficiently, both from a productivity and 
a performance point of view, arises. Explicit parallel programming is hard and automatic 
means of parallelisation, while giving a sequential programming model to the programmer, 
often misses the target when it comes to performance. {\bf \emph{source??}}

@@

\subsection{GPU programming}

@@

There are some superficial differences between the code generated using Obsidian and the 
handwritten example above. One such difference is that the index computation is not shared. 
Obsidian relies on the CUDA compiler to discover such sharing via a common subexpression 
elimination optimisation pass. Another difference is that Obsidian has inserted a conditional 
that only executes the body for certain blocks. Discussion of the causes for that conditional 
is postponed until a later section {\bf \emph{POINT TO SECTION}}. A real difference is that 
the kernel is generated for a fixed size, here there is a constant (32) where in the handwritten
code we used {\tt blockDim.x}. If those differences are set aside, the two kernels are identical. 
Figure~\ref{fig:OBSIDIANHOST1} shows how the {\tt vecAddG} kernel can be launched from 
within Haskell, using Obsidian's rudimentary support for launching GPU computations.

@@

In Obsidian, local computations can also use shared memory. The Obsidian arrays 
we have seen in the examples, {\tt Pull} and {\tt Push} arrays do not directly 
correspond to actual data in memory. Pull and push arrays, rather represent ways 
to compute arrays. More information about these array representations is available 
in section {\bf \emph{REF SECTION}}. The Obsidian programmer can store intermediate 
arrays in shared memory by using a {\tt force} operation. The 
{\tt force} operation computes an array and stores it into shared memory.

@@

\noindent This kernel computes its result without any storage of intermediate results. 
In the generated CUDA code below, it can be seen that the (*2) and (+1) operations 
are applied without any intermediate storage.

@@

@@

In Obsidian it is impossible for a barrier synchronisation to appear within diverging 
conditional code. Conditionals can be introduced either by an {\tt ifThenElse} function 
at the element level or by a {\tt Cond} statement that is an operation in the {\tt Program} 
monad. The {\tt Program} monad is parameterised on the GPU hierarchy level. There are 
thread programs ({\tt TProgram}), block programs ({\tt BProgram}) and grid programs ({\tt GProgram}). the {\tt Cond} conditional choose between two thread programs while a synchronisation can 
only appear in an Obsidian block program. More information about programs and the GPU hierarchy 
is available in section~\ref{sec:OBSIMP}.

@@

{\bf {\emph old text from here to end}}

% ---------------------------------------------------------------------------
% ---------------------------------------------------------------------------
%
% Embedded languages
%
% ---------------------------------------------------------------------------
% ---------------------------------------------------------------------------
\section{Embedded languages}

This thesis applies embedded language methodology to data-parallel programming. An 
embedded language is implemented as a library in a host language,  Haskell in this case. 
The embedded language approach is beneficial since we are not entirely sure how 
to program current and future highly parallel computers. Implementing embedded languages allows 
us more rapid prototyping of ideas. Other benefits of embedded languages often listed 
include making use of the host language's type system to guarantee that correctly typed target code is generated. 
Already existing libraries in the host language (if they are general enough) 
can be used in the embedded language. The host language can also be used as a powerful macro 
language for your embedded language.

A very good paper, and highly relevant to our approach, is ``Compiling Embedded Languages''
by Conal Elliott et al~\citet{COMPILEEDSL}. This paper describes how to embed a compiler 
backend and generate efficient code in some target language. One key aspect of this 
approach is that the embedded language is a library of functions that create and 
compose Abstract Syntax Trees (ASTs). Obsidian (chapter \ref{chap:GPUProgramming}) as 
well as EmbArBB (chapter \ref{chap:ArBB}) are implemented in a similar way. An 
embedding that builds ASTs is called a deep embedding. Shallow embeddings, on the 
other hand, do not build ASTs. In ``Combining Deep and Shallow Embedding for EDSL'' 
Josef Svenningsson and Emil Axelsson combine the two methods, shallow and deep, 
in order to simplify the AST data type (fewer constructors) and make more extensible 
embedded languages \citet{DEEPSHALLOW}. Obsidian also uses a combination of shallow 
and deep embeddings, while EmbArBB is a more traditional deep embedding.

@@

% ---------------------------------------------------------------------------
% ---------------------------------------------------------------------------
%
% Obsidian
%
% ---------------------------------------------------------------------------
% ---------------------------------------------------------------------------
\section{Obsidian: An embedded language for GPU kernel implementation}

\vspace{5mm} 
\emph{This section will be revised. I want to add more code examples.} 
\vspace{5mm}

Obsidian aspires to raise the level of abstraction of GPU kernel implementation, 
while maintaining enough control over details that influence performance on a GPU. 
We try to reach these goals by offering abstractions to the programmer that are 
compositional and parallel. We also ensure that these can easily be compiled 
to NVIDIA CUDA code. We have two different array representations, each with 
its strengths and weaknesses. We also feel that it is important that the programmer 
be able to use local shared memory in computations and have the ability to influence 
memory access patterns. What these array representations and abstractions are 
will be explained in the following sections, with references to the papers 
that introduce them or explain them the best.

\subsection{Deep embedding: The {\tt Program} data type}

@@

\subsection{Scalars}

@@

\subsection{Arrays}

@@

\subsubsection{Pull arrays}

@@

Pull arrays are parameterised on both element ({\tt a}) and length ({\tt s}) type. The length 
parameter is used to be able to represent arrays with both static (known at Haskell runtime) 
or dynamic (length represented by an expression, an unknown length). When creating block and 
warp level computations, we want to know the exact lengths in order to avoid generating code 
riddled with conditionals concerning array lengths. It is possible to write kernels that 
are not specialised for a certain size. This is accomplished by adding bounds checks. Since 
conditionals  are problematic from a performance point of view, fixed size kernels are recommended 
when performance is crucial. 
%Conditionals are problematic on GPUs and 
%while general programs that run for any size is possible when performance is crucial a fixed 
%size is recommended.

@@

\subsubsection{Push arrays}

@@

\subsection{Compilation to CUDA}

@@

\emph{This list below will be revised. More details or code will be added.}

\item {\bf Minimise data transfers between host and device:} Data transfers
between host and device are out of Obsidian's scope. If Obsidian is used to create 
a kernel for a computation, the programmer has already decided that a GPU should be used.

\item {\bf Ensure that global memory accesses are coalesced:} Inputs to an Obsidian 
program are pull arrays. Pull arrays can be arbitrarily permuted; the permutation 
used will decide the access pattern. If this method can be applied or not partly also 
depends on the algorithm being implemented.

\item {\bf Minimise use of global memory:} Intermediate results can be computed into 
shared memory. Arrays are stored into shared memory when the programmer uses a 
{\tt force} primitive.

\item {\bf Use a multiple of 32 threads per block:} Obsidian requires regularity even more 
than CUDA does. Since we focus on generating high performance kernels, we rule out generation 
of kernels that work on different input sizes. The choise of how many threads to use per block 
is in the hands of the Obsidian program. The method for making that choise has changed over time. 
Earlier, the number of threads needed by kernel was decided by the size of the arrays it 
computed. This has recently changed is now set by parameters to the code generation process.

\item {\bf Avoid different execution paths within a block:} Sometimes divergent code cannot 
be avoided. However, It is important not to have threads diverge unnecessarily. When using pull 
arrays alone this was often a problem; with the addition of push arrays the situation improved. 
Using push arrays a divergent computation can be explicitly split up into phases, each with all 
threads following the same execution path. This means that fewer threads are required to do the 
same work.

\item {\bf Do not use {\tt \_\_synchthreads()} in diverging code:} In Obsidian it is impossible 
for a barrier synchronisation to appear within conditional code. Conditionals can be introduced 
either by a conditional expression or a {\tt Cond} operation in the program datatype. The Cond 
operation chooses between two thread programs and the synchronisation operation is at the block 
level in the hierarchy. So a synchronisation cannot occur in conditional code. 
%f it is safe or 
%not to call some function at a certain place in your code, can only be decided by inspecting 
%that functions implementation. 
\end{itemize}

%% In the book ``Structured Parallel Programming''~\citet{STRUCTURED}, there is a list showing language 
%% support for various parallel patterns. In figure~\ref{fig:caps} this list augmented with a column for Obsidian. 
%% This figure is added to indicate relative expressive power of the different languages. The most 
%% interesting comparison is between Obsidian and OpenCL that both have a programming model geared towards
%% GPUs; even though OpenCL can compile for both CPU and GPU execution. The paragraphs below 
%% take a closer look at well Obsidian performes in the patterns.

%% \subsubsection{Map}

%% Map, to a  functional programmer is a higher order function that takes a pure 
%% function and applies it to all elements of a collection (for example a list). 
%% That meaning of map is easily implementable in Obsidian on both pull and push arrays. 
%% Obsidian is also capable, via its parallel for loops, to map sequential computations 
%% over threads in a block and block computation over many blocks. This is probably 
%% more in the spirit of what the authors refer to in~\citet{STRUCTURED}.

%% \subsubsection{Stencil}

%% Stencil computations can be expressed in Obsidian, with some manual labour. 
%% A stencil computation would have the form of a pull array to pull array function. 
%% Pull arrays allow reading at position. 
%% Stencil operations where more than one output position is touched could be 
%% implemented as a pull array to push array function.
%% One problem with stencil computations is the borders. If every thread needs to 
%% execute a conditional, checking if it is at the border, performance would suffer. 
%% Using Obsidian the programmer would likely generate more than one kernel; a fast 
%% kernel for the internal points and one or more border kernel.

%% \subsubsection{Reduction and scan} 
%% Reductions can easily be implemented in Obsidian and with very good performance. 
%% Obsidian also seems particularly well suited for experimenting with the kind 
%% of trade offs that are interesting when implementing reduction kernels.

%% Scans can also be implemented and the code is elegant.

%% \subsubsection{Pack and expand}

%% \begin{figure}[!ht]
%% \begin{small} 
%% \begin{tabular}{|l|l|l|l|l|l|l|}
%% Parallel Pattern & TBB & Cilk Plus & OpenMP & ArBB & OpenCL & Obsidian \\
%% \hline 
%% Parallel Nesting & F   & F         &        &      &        &          \\
%% Map              & F   & F         & F      & F    & F      & I        \\
%% Stencil          & I   & I         & I      & F    & I      & I        \\ 
%% Workpile         & F   &           &        &      & I      &          \\
%% Reduction        & F   & F         & F      & F    & I      & I        \\ 
%% Scan             & F   & I         & I      & F    & I      & I        \\ 
%% Fork-Join        & F   & F         & I      &      &        &          \\ 
%% Recurrance       &     & P         &        &      &        &          \\ 
%% Superscalar seq. &     &           &        &      & F      &          \\

%% Pack             & I   &  I        &  I     & F    & I      & I        \\ 
%% Expand           & I   &  I        &  I     & I    & I      & *        \\
%% Pipeline         & F   &  I        &  I     &      &        &          \\
%% Geometric decomp. & I  &  I        &  I     & I    & I      & I        \\
%% Search           &  I  &  I        &  I     & I    & I      & I        \\
%% Category reduction & I &  I        &  I     & I    & I      & I        \\
%% Gather           & I   &  F        &  I     & F    & I      & I        \\
%% Atomic scatter   & F   &  I        &  I     &      & I      & I        \\
%% Permutation scatter & F & F        &  F     & F    & F      & I        \\ 
%% Merge scatter    &  I  &  I        &  I     & F    & I      & *        \\
%% Priority scatter &     &           &        &      &        & *        \\

%% \end{tabular}
%% \end{small}
%% \label{fig:caps}
%% \caption{This figure lists what parallel patterns a number of languages support. The list 
%% is copied from ``Structured Parallel Programming''~{\protect\citet{STRUCTURED}} and augmented with a 
%% column for Obsidian. \\
%% {\bf Key:} F - means the pattern is supported directly with an abstraction. P - indicates 
%% that the pattern can be implemented in terms of other patterns appearing below. I - means 
%% that the pattern can be easily implemented. Lastly, (*) indicates some patterns 
%% that can be implemented in Obsidian but currently not efficiently so.} 
%% \end{figure}

%% In the technical report ``The Landscape of Parallel Computing Research: A View 
%% from Berkeley''~\citet{dwarfs}, 13 computational kernels (called dwarfs) of importance is identified. 
%% The paper also proposes that we use these 13 dwarfs to evaluate parallel programming 
%% models and languages. The dwarfs are listed below, together with thoughts on how 
%% well they can be tackled by Obsdian.

%% \begin{itemize} 
%% \item {\bf Dense Linear Algebra:} Obsidian works well for expressing vector-vector 
%% operations. It is also possible to express operations such as matrix multiplication, 
%% even though Obsidians support for multidimensional arrays is limited. It is possible 
%% to use pull arrays of pull arrays, but not push arrays of push arrays.

%% \item {\bf Sparse Linear Algebra:}

%% \item {\bf Spectral Methods (FFT):} FFTs are implemented in a way that is 
%% similar to sorting networks. A so-called butterfly network is used. We know 
%% that we can express these network structures in Obsidian.

%% \item {\bf N-Body Methods:}

%% \item {\bf Structured Grids:} Regular grids should not be a problem for Obsidian. 
%% Obsidian has no built in support for stencils, the combination of pull and push array 
%% still provides the capability to express such computations.

%% \item {\bf Unstructured Grids:} May be hard problem for GPUs in general. The paper 
%% states that these problems have the property that ``Updates typically involve multiple 
%% levels of memory reference indirection''. We know that the GPU thrives when memory access 
%% patters are predictable and regular, which is the opposite scenario.

%% \end{itemize}

@@

% ---------------------------------------------------------------------------
% ArBB EmbArBB
% ---------------------------------------------------------------------------
The ArBB and EmbArBB systems also provide a set of parallel primitives that the programmer
composes to build his application. I place the ArBB and EmbArBB in a box slightly lower than 
Accelerate since while ArBB/EmbArBB also have built in reduction primitives, just like Accelerate, 
they are not general primitives. Where Accelerate has a single higher order reduction 
operation, ArBB/EmbArBB has {\tt reduce\_add, reduce\_mul} and so on. Benefits of 
having ArBB embedded in Haskell compared to C++ are discussed in section~\ref{sec:EmbArBB}.
%I will go into 
%what benefits there are to having ArBB embedded in Haskell compared to C++ in

@@

Obsidian is a language for GPU kernel implementation. It is not as well integrated 
into Haskell as for example Accelerate. There are ways to execute Obsidian kernels 
from within Haskell; the recommended workflow, however, is to generate the CUDA  
as text and write the host code by hand. This can be seen as both a shortcoming 
and a feature. It is hard to integrate Accelerate programs in larger applications 
written in other languages; when using Obsidian a kernel is generated and 
can be used as part of for example a library.

@@

Here and there in the thesis it has been hinted that Obsidian has warp level 
capabilites and that these are not as integrated into the hierarchy as the block, 
thread and grid levels are. The warp abstractions offered also requires more work in 
the code generating backend than the others, that correspond directly to CUDA concepts. 
The problem is that CUDA does not have a warp abstraction. Warps needs to be 
better integrated with the rest of Obsidian and the code generation required 
for warps needs more testing before being trustworthy and useful.

@@

\emph{This section will revised.}

This thesis contains two approaches to embedding a language for data-parallel programming 
in Haskell. EmbArBB embeds a preexisting library for data-parallelism in Haskell 
and at little effort from us, very good performance is gained. The performance of 
EmbArBB is entirely due to all the optimisation expertice and effort that was invested in 
the ArBB system. We get all that expertice for free. What EmbArBB provides is a 
way for Haskellers to benefit from the ArBB system. ArBB and EmbArBB are based on a 
set of parallel primitives for the programmer to use and compose. Obsidian takes 
a different approach. Built in primitives are very few and very low level, such as 
sequential and parallel for loops. On top of the low-level capabilities, a library 
of high-level primitives can be built. The key is that we want the programmer to 
be able to specify how to implement these higher level abstractions.

@@

Performance of Obsidian kernels is often good and the Obsidian code that they 
are generated from is often shorter than corresponding CUDA programs. Obsidian 
programs differ from CUDA programs by describing whole programs. With this I mean 
that a CUDA programmer needs to take a leap from the single threaded program 
parameterised on thread Id, to what actually will happen when this program is 
run by hundreds or thousands threads at once. The extensive indexing arithmetic 
often present in CUDA programs can make this leap a hard one. Similar programs 
in Obsidian could use operations on whole arrays such as {\tt map} or explicitly 
bounded parallel loops.

>>>>

% ---------------------------------------------------------------------------
% ABSTRACT !!  
% ---------------------------------------------------------------------------
\section*{Abstract}

Computers today are becoming more and more parallel. General purpose processors (CPUs) 
have multiple processing cores and Single Instruction Multiple Data (SIMD) units for 
data-parallelism. Graphics processors (GPUs) brings massive parallelism at the cost of 
being harder to program than CPUs. This thesis applies embedded language methodology 
to data-parallel programming. Two embedded languages are presented, Obsidian for general 
purpose GPU programming and EmbArBB for data-parallel programming across platforms.

Each generation of CPU and GPU bringing more parallel resources. The question of how to write 
programs for these processors efficiently arise. We are after efficiency both in programmer 
productivity and in application performance. Using embedded languages allows us to experiment
with what abstractions to present to the programmer at relatively little effort.

Obsidian is an embedded language for general purpose programming of GPUs. We try to strike 
a balance between high level, productivity increasing abstractions and low-level 
control needed for performance. Obsidian programs are compiled into CUDA C code.

In cases where the programmer is not interested in low-level details 
or wants the program to run over a range of hardware a higher level language can be used.
EmbArBB is a Haskell embedding or the Intel ArBB system. EmbArBB relies on the ArBB system 
to generate code (via a Just-In-Time compiler) to a range of hardware.

EmbArBB embeds a preexisting library for data-parallelism into Haskell and we obtain 
very good performance at little implementation effort. This performance comes from the 
expertise and effort put into the ArBB system and that we get for free. Embedding ArBB is 
a way to provide these benefits to the Haskell programmer and a way to increase usefulness
of an existing system by opening it up to a wider audience. Obsidian is very different, it 
is not based on a set of high-level parallel primitives. The Obsidian programmer 
can implement these primitives in different ways, select the best one and then  abstract 
upon that. We have obtained very good performance in case studies involving reductions.

@@

I would like to thank my supervisor Mary Sheeran. She is a very suportive, motivating and 
understanding supervisor. I also thank my co-supervisor Koen Claessen for all the great advice 
and tips he has provided over he years. I cannot imagine that this would have worked with 
any other supervisors. Thanks to Josef Svenningsson, a late addition to my troop of 
supervisors. Working with Josef has been a very fun and rewarding experience.

I thank Manuel Chakravarty for letting me visit his group at the University of New South Wales 
and Trevor McDonell and Sean Lee for taking care of me during my visit. I look back 
very fondly to my stay in Australia.

For being a great supervisor during my three month internship at Intel, I thank Ryan Newton. 
These where three very intense months away from family but at the same time a very rewarding 
experience.

I want to thank Erik, Ola, Markus and Viktor for our occasional lunches and GPU discussions.

Thanks to my thesis opponent Stephen Edwards for valuable comments on a draft of this thesis.

Finally I want to thank my family and friends. Thanks to my sisters and parents for their support.
I thank my friends for being understanding during times when I have been very self-absorbed. 
I am very grateful to my wife Andrea for supporting and believing in me and our daughter Noemi 
for bringing much joy to my breaks from work while writing this thesis; soon she is old enough 
for real LEGO.

@@

Computers today are increasingly parallel. Each new generation of general purpose processors
(CPUs) and of graphics processors (GPUs), brings more parallelism. The question 
of how to program these computers most efficiently, both from a productivity and 
a performance point of view, arises. Explicit parallel programming is hard and automatic 
means of parallelisation, while giving a sequential programming model to the programmer, 
often misses the target when it comes to performance. The book ``Structured Parallel 
Programming'' by Michael McCool et al argues for the need of explicit parallel programming 
models~\citet{STRUCTURED}.

@@

\subsection{GPU programming} 
\label{GPUPROGRAMMING}

@@

There are some superficial differences between the code generated using Obsidian and the 
handwritten example above. One such difference is that the index computation is not shared. 
Obsidian relies on the CUDA compiler to discover such sharing via a Common Subexpression 
Elimination (CSE) optimisation pass. Another difference is that Obsidian has inserted a 
conditional that only executes the body for certain blocks. Discussion of the causes for 
that conditional is postponed until a later section~\ref{sec:OBSARRAYS}. A real 
difference is that the kernel is generated for a fixed size, here there is a constant (32) 
where in the handwritten code we used {\tt blockDim.x}. If those differences are set 
aside, the two kernels are identical. Figure~\ref{fig:OBSIDIANHOST1} shows how the 
{\tt vecAddG} kernel can be launched from within Haskell, using Obsidian's rudimentary 
support for launching GPU computations.

@@

In Obsidian, local computations can also use shared memory. The Obsidian arrays 
we have seen in the examples, {\tt Pull} and {\tt Push} arrays do not directly 
correspond to actual data in memory. Pull and push arrays, rather represent ways 
to compute arrays. More information about these array representations is available 
in section~\ref{sec:OBSARRAYS}. The Obsidian programmer can store intermediate 
arrays in shared memory by using a {\tt force} operation. 
%The 
%{\tt force} operation computes an array and stores it into shared memory.

@@

\noindent This kernel computes its result without any storage of intermediate results;
the code is equivalent to \verb!fmap ((+1).(*2))!. 
In the generated CUDA code below, it can be seen that the (*2) and (+1) operations 
are applied without any intermediate storage.

@@

The code below is generated using the {\tt kernelF} program:

\begin{small} 
\begin{Verbatim}[samepage=true]
extern "C" __global__ void kernel(int32_t* input0, uint32_t n0,
                                  int32_t* output1)
{
    extern __shared__ uint8_t sbase[];
    uint32_t tid = threadIdx.x;

    if (blockIdx.x < n0 / 32U) {
        ((int32_t*) sbase)[tid] = input0[blockIdx.x * 32U + tid] * 2;
        __syncthreads();
        output1[blockIdx.x * 32U + tid] = ((int32_t*) sbase)[tid] + 1;
    }
}
\end{Verbatim}
\end{small}

\noindent This code uses a shared memory array called {\tt sbase}. There 
is also a call to {\tt \_\_syncthreads} to ensure that all the writes have 
completed before proceeding. In this case the synchronisation is not really
necessary since all threads belong to the same warp. There is a {\tt unsafeforce} 
operation that naively removes synchronisations if the array being forced is short 
enough. More advanced analysis could also be performed to remove synchronisations 
whenever communication is entirely warp local. We do not apply that optimisation, 
but work on warp level computations is in progress and would put this ability 
in the hands of the programmer.

@@

In Obsidian it is impossible for a barrier synchronisation to appear within diverging 
conditional code. Conditionals can be introduced either by an \\{\tt ifThenElse} function 
at the element level or by a {\tt Cond} statement that is an operation in the {\tt Program} 
monad; we have seen an example of the program monad in section~\ref{sec:GPUPROGRAMMING}. 
The {\tt Program} monad is parameterised on the GPU hierarchy level. There are 
thread programs ({\tt TProgram}), block programs ({\tt BProgram}) and grid programs 
({\tt GProgram}). the {\tt Cond} conditional choose between two thread programs while a 
synchronisation can only appear in an Obsidian block program. More information about programs 
and the GPU hierarchy is available in section~\ref{sec:OBSIMP}.

@@

Obsidian is implemented as an embedded language using Haskell as host. Obsidian 
is a compiled embedded language; a very good paper that is highly relevant to our 
approach, is ``Compiling Embedded Languages'' by Conal Elliott et al~\citet{COMPILEEDSL}. 
This paper describes how to embed a compiler backend and generate efficient code 
in some target language. One key aspect of this approach is that the embedded 
language is a library of functions that create and compose Abstract Syntax Trees 
(ASTs). Obsidian (chapter \ref{chap:GPUProgramming}) as well as EmbArBB 
(chapter \ref{chap:ArBB}) are implemented in a similar way. An embedding that builds 
ASTs is called a deep embedding. Shallow embeddings, on the other hand, do not build 
ASTs. In ``Combining Deep and Shallow Embedding for EDSL'' Josef Svenningsson and 
Emil Axelsson combine the two methods, shallow and deep, in order to simplify the 
AST data type (fewer constructors) and make more extensible embedded languages 
\citet{DEEPSHALLOW}. Obsidian also uses a combination of shallow 
and deep embeddings.

@@

\subsubsection{Deep embedding: The {\tt Program} data type}

@@

\subsubsection{Scalars}

@@

\subsubsection{Arrays} 
\label{sec:OBSARRAYS}

@@

\paragraph{Pull arrays:}

@@

Pull arrays are parameterised on both element ({\tt a}) and length ({\tt s}) type. The length 
parameter is used to be able to represent arrays with both static (known at Haskell runtime) 
or dynamic (length represented by an expression, an unknown length). When creating block 
level computations, we want to know the exact lengths in order to avoid generating code 
riddled with conditionals concerning array lengths. It is possible to write kernels that 
are not specialised for a certain size. This is accomplished by adding bounds checks. Since 
conditionals  are problematic from a performance point of view, fixed size kernels are recommended 
when performance is crucial. Dynamic lengths are used for grid level computations, this allows 
a local computation of a fixed size to be run in parallel over arrays a multiple of the size that 
the local computation handles. In generated code this is currently visible as a conditional 
on {\tt blockIdx.x}. However, this is still considered work in progress; the conditional is 
only really necessary for Obsidian kernels that have more than one output array and the outputs
are of different length. 
%Conditionals are problematic on GPUs and 
%while general programs that run for any size is possible when performance is crucial a fixed 
%size is recommended.

@@

\paragraph{Push arrays:}

@@

\subsubsection{Compilation to CUDA}

@@

@@

% ---------------------------------------------------------------------------
% ArBB EmbArBB
% ---------------------------------------------------------------------------
The ArBB and EmbArBB systems also provide a set of parallel primitives that the programmer
composes to build his application. I place the ArBB and EmbArBB in a box slightly lower than 
Accelerate since while ArBB/EmbArBB also have built in reduction primitives, just like Accelerate, 
they are not general primitives. Where Accelerate has a single higher order reduction 
operation, ArBB/EmbArBB has \\ {\tt reduce\_add, reduce\_mul} and so on. Benefits of 
having ArBB embedded in Haskell compared to C++ are discussed in section~\ref{sec:EmbArBB}.
%I will go into 
%what benefits there are to having ArBB embedded in Haskell compared to C++ in

@@

Obsidian is a language for GPU kernel implementation, not complete applications. 
It is not as well integrated into Haskell as for example Accelerate. There are 
ways to execute Obsidian kernels from within Haskell (we have seen examples of this in 
section~\ref{sec:GPUPROGRAMMING}); the recommended workflow, 
however, is to generate the CUDA as text and write the host code by hand. This 
can be seen as both a shortcoming and a feature. It is hard to integrate 
Accelerate programs in larger applications written in other languages; when 
using Obsidian a kernel is generated and can be used as part of for example a library.

@@

Warp level computations is work in progress and that these are not as well integrated 
into the hierarchy as the block, thread and grid levels are. The warp abstractions 
offered also requires more work in the code generating backend than the others, 
that correspond directly to CUDA concepts. The problem is that CUDA does not have 
a warp abstraction. Warps needs to be better integrated with the rest of Obsidian 
and the code generation required for warps needs more testing before being 
trustworthy and useful. When operational, warp level abstractions will help 
the programmer make better use of the GPU.

@@

%\vspace{5mm}

% \emph{This section will revised.}

%\vspace{5mm}

This thesis contains two approaches to embedding a language for data-parallel programming 
in Haskell. EmbArBB embeds a preexisting library for data-parallelism in Haskell 
and at little effort from us, very good performance is gained. The performance of 
EmbArBB is entirely due to all the optimisation expertise and effort that was invested in 
the ArBB system. We get all that expertice for free. What EmbArBB provides is a 
way for Haskellers to benefit from the ArBB system. ArBB and EmbArBB are based on a 
set of parallel primitives for the programmer to use and compose. Obsidian takes 
a different approach. Built in primitives are very few and very low level, such as 
sequential and parallel for loops. On top of the low-level capabilities, a library 
of high-level primitives can be built. The key is that we want the programmer to 
be able to specify how to implement these higher level abstractions.

@@

Performance of Obsidian kernels is often good and the Obsidian code that they 
are generated from is often shorter than corresponding CUDA programs. Obsidian 
programs differ from CUDA programs by describing whole programs. With this I mean 
that a CUDA programmer needs to take a leap from the single threaded program 
parameterised on thread ID, to what actually will happen when this program is 
run by hundreds or thousands threads at once. The extensive indexing arithmetic 
often present in CUDA programs can make this leap a hard one. Similar programs 
in Obsidian could use operations on whole arrays such as {\tt map} or explicitly 
bounded parallel loops.
